\textbf{ Spectrophotometric parallaxes with linear models:\\ Accurate distances for luminous red-giant stars }

### 2018-10-14 paper/sp.tex
hogg edits

% To-do items
% -----------
% - Coryn's point that we should carefully explain how our model is geometric.
% - check with WISE experts that we have the correct WISE photometry
% - write words about other linear models we have made or could make, probably in the discusssion.
% - are we citing all the proper DR2 papers we need to cite?
% - HWR: Footnote to explain the use of the word ``feature'' for astronomers.
% - HWR: Figures 1 and 2 need to be combined to directly compare Parent and Training.
% - HWR: Figure 3: Note that the y-axis is positive-definite!
% - HWR: Figure 7: Caption needs to say what arrows represent.
% - Make the comparison with Cannon as a function of L1 strength and maybe look at correlation coefficients between parameter vectors.
% - Get model running on ASPCAP outputs and compare in Discussion.
% - Make sure we incorporate / respond to Ness comments from Aug 13 email. And thank Ness.
% - Thank APOGEE telecon and add paragraph about AGB stars and the color-magnitude diagram in the discussion.
% - Describe how we ran The Cannon; discussion of how we compare to The Cannon.
% - Cite and describe Eilers et al.
% - comment on parallax offset being a function of position and color
% - remove Cannon plot for the moment
% - comment on new plot of new precision on parallax compared to Gaia's precision
% - Get figure showing how much better we are than Gaia at large distance and as a function of distance. And write a paragraph about said figure.

@@

% math macros
\newcommand{\T}{^{\mathsf{T}}}
\DeclareMathOperator*{\argmin}{argmin}
\newcommand{\logg}{\log g}
\newcommand{\BP}{{G_\mathrm{BP}}}
\newcommand{\RP}{{G_\mathrm{RP}}}
\newcommand{\gparallax}{\varpi^{(\mathrm{a})}}
\newcommand{\sparallax}{\varpi^{(\mathrm{sp})}}

@@

\begin{abstract}\noindent
% context
With contemporary infrared spectroscopic surveys like \apogee,
red-giant stars can be observed to distances and extinctions
at which \gaia\ parallaxes are not informative.
Yet the combination of effective temperature, surface gravity, composition, and age---all
accessible through spectroscopy---determines a giant's luminosity.
Therefore spectroscopy plus photometry should enable 
precise spectrophotometric distance estimates.
% aims
Here we use the \apogee--\gaia\ overlap to train a data-driven model
to predict parallaxes for red-giant branch stars with $0<\logg\leq2.2$.
% methods
We employ (the exponentiation of)
a linear function of \apogee\ spectral pixel intensities and multi-band photometry
to predict a parallax spectrophotometrically.
The model training involves no logarithms or inverses of the \gaia\ parallaxes,
and needs no cut on the \gaia\ parallax signal-to-noise ratio.
It includes an L1 regularization to zero out the contributions of
uninformative pixels.
The training is performed
with leave-out subsamples such that no star's astrometry is used even indirectly in its
spectrophotometric parallax estimate.
% results
The model implicitly performs a reddening and extinction correction in its parallax prediction,
without any explicit dust model.
We assign to each star in the sample a new spectrophotometric parallax estimate;
these parallaxes have uncertainties of a few to 15 percent, depending on data quality,
which is more precise than the \gaia\ parallax for the vast majority of targets.
We can now get 10-percent distance estimates out to heliocentric distances of 20\,kpc,
allowing truly global maps of the Milky Way's disk.
\end{abstract}

@@

If we want to make precise kinematic and element-abundance maps
of the Milky Way disk out to large heliocentric radii,
and especially through the extinction to the other side of the Galactic Center, we will
need to make use of luminous red giants, and we will need to observe in
the infrared.
These arguments motivate the \apogee\ (\citealt{apogee}),
\apogee-2, and \sdssv\ (\citealt{sdssv})
projects, which take spectra in the infrared, and deliver detailed abundances
along the entire red-giant branch up to the tip.
The maps made with these data will reveal the dynamics of the Milky Way and its disk,
and show us how and where stars form, and how they migrate around (and out of) the
disk with time.

@@

Luminous red-giant stars are valuable for mapping the Milky Way disk for a number
lf reasons, only one of which is their great luminosities (and hence brightnesses
even at large distance).
They are extremely common (frequent) stars, much more common than distance-indicating
stars like Cepheids.
They are produced by stellar populations at all metallicities and almost all ages,
so they can be found in all parts of the Galaxy.
They have temperatures and surface gravities such that it is
straightforward to spectroscopically measure metallicities,
detailed chemical abundances, and stellar parameters, even including ages (\citealt{martig, nessage}).
The red-giant branch is photometrically near-orthogonal to reddening vectors by dust,
so they are easy to de-redden or dust-correct.
And finally, in physical models of red giants, their predicted luminosities are simple
functions of composition, surface gravity, temperature, and age.
Thus, if we can measure these properties of red giants, we can (in principle) predict
their luminosities and use them as standard candles.
That is---like type Ia supernovae---red giants are standardizable candles.

@@

Data-driven models are very often more accurate than physical models of stars
for this kind of work, because they have the flexibility to discover
regularities or offsets in the data that are not captured by the physical
models.
They do better than physical models when there are lots of training data
with properties that are good for performing regressions.

@@

\item[linearity] Perhaps the most restrictive assumption of this work is that
the logarithmic parallax (or, equivalently, the distance modulus) 
can be predicted as a \emph{completely linear function} of
the chosen features. We are only assuming this linearity in a small range of stellar
surface gravity, but this assumption is strong, and limits strongly the
flexibility or capacity of the model.
We make it to ensure that our method is easy to optimize, and the results are easy
to interpret.

\item[likelihood] We assume that the \gaia\ parallaxes are generated by a particular
stochastic process, in which the difference between the \gaia-Catalog parallax (plus
a small \gaia-recommended offset; \citealt{lindegren}) and the true parallax is effectively drawn from a
Gaussian with a width set by the \gaia-Catalog uncertainty on the parallax.
This is the standard assumption in all properly probabilistic \gaia-Mission inferences
to date, but it subsumes a number of related assumptions, like that the \gaia\ noise
model is correct, that the stars are only covariant with one another at negligible levels, and that
there are no significant outliers or large-scale systematics.
\end{description}

@@

The model and log-likelihood function can be expressed heuristically as
\begin{eqnarray}
\gparallax_n &=& \exp(\theta\cdot x_n) + \mbox{noise}
\\
\chi^2(\theta) &\equiv& \sum_{n=1}^N \frac{[\gparallax_n - \exp(\theta\cdot x_n)]^2}{\sigma_n^2}
\end{eqnarray}
where
$\gparallax_n$ is the \gaia\ measurement (adjusted; see below) of the astrometric parallax of star $n$,
the model is that the logarithm of the true parallax
can be expressed as a linear combination of the components
of some $D$-dimensional feature vector $x_n$,
$\theta$ is a $D$-vector of linear coefficients,
$\sigma_n$ is the \gaia\ estimate of the uncertainty on the parallax measurement,
and $\chi^2(\theta)$ is (twice) the negative-log-likelihood for the parameters $\theta$
under the assumption of known Gaussian noise and
that there are $N$ independently measured stars $n$.
The feature vector $x_n$ contains photometry in a few bands, and all 7400-ish pixels
in the pseudo-continuum-normalized \apogee\ spectrum, so $D$ is on the order of 7400.

In addition, we assume that many entries in the parameter $D$-vector $\theta$ will be zero
(the ``sparsity'' assumption).
In order to encourage this, we optimize not $\chi^2$ but a regularized objective function
\begin{eqnarray}
\hat{\theta} &\leftarrow& \argmin_{\theta}\left[\frac{1}{2}\,\chi^2(\theta) + \lambda\,||P\cdot\theta||_1^1\right]
\label{eq:objective}
\end{eqnarray}
where
$\lambda$ is a regularization parameter,
$P$ is a projection operator that selects out from the $\theta$ vector only those components
that pertain to the \apogee\ spectral pixels,
and $||y||_1^1$ is the L1-norm or sum of absolute values of the components of $y$.
This kind of regularization (L1) adds a convex term to the optimization and leads to
sparse optima.
The $P$ operator makes it such that we only regularize the components of $\theta$ that multiply
the spectral pixels; we ask for the spectral model to be sparse but we don't ask for the photometric
model to be sparse.
We set the value of $\lambda$ by cross-validation.

@@

Once the model is optimized, the output of the model is a prediction of the parallax,
or really what we will call the \emph{spectrophotometric parallax}.
This spectrophotometric parallax
$\sparallax_m$ for any star $m$ in the validation or test set is
assigned according to
\begin{eqnarray}
\sparallax_m &\leftarrow& \exp(\hat{\theta}\cdot x_m)
\end{eqnarray}
where
$\hat{\theta}$ is the optimal parameter vector according
to \equationname~(\ref{eq:objective}),
and
$x_m$ is the feature $D$-vector for star $m$.
Thus the model can be trained on a training set of stars and applied to
a validation set or a test set to make predictions.
The only requirements are that every test-set object $m$  must have a full feature
vector $x_m$ just as every training-set object $n$ must have a full feature
vector $x_n$.

@@

We used the \gaia\ (\citealt{gaia}) \acronym{DR2} (\citealt{gaiadr2}) Data Archive (\citealt{gaiaarchive})
official matches (\citealt{xmatch}) to match the \apogee+\zmass\ stars
to the \wise\ Catalog (\citealt{wise}),
according to the \gaia\ Archive internal match criteria.
This gives, fore each matching star $n$,
mid-infrared photometry $W_{1n}$ and $W_{2n}$ at 3.6 and 4.5\,$\mu$m
In detail we use the \code{w1mpro} and \code{w2mpro} Catalog entries.

@@

For training, we need A and B Training Sets.
We define the Training Set stars to be Parent-Sample stars that also
have a measured \gaia\ parallax.
The Training-Set stars, in addition to meeting all Parent-Sample cuts,
also must meet $\sigma_{\varpi^{(\rm a)}} < 0.1$, it must be observed by
more than $8$ (widely separated) times
\begin{equation}
\code{visibility_periods_used >= 8}
\end{equation}
as well as meet the additional quality criterion
\begin{equation}
\code{astrometric_chi2_al / sqrt(astrometric_n_good_obs_al - 5) <= 35}
\end{equation}
which eliminates detections that are not consistent with a single PSF \citep[see][]{calj},
to ensure that the parallax measurements are good.
These cuts leave 28226 stars total in the Training Sets.
But---as we have emphasized above---we do not cut ever on parallax or
parallax error or their ratio. For this reason, most of the Training-Set stars
do not have significantly measured parallaxes and more than 2~percent even have
negative parallaxes!
But we will perform our training such that it will be unbiased in these
circumstances.
Indeed it would be necessarily biased if ever we did cut on parallax or
parallax signal-to-noise.

In detail,
for each star $n$ in the Training Sets, we take from \gaia\ \acronym{DR2} (\citealt{gaiadr2})
the astrometric parallax $\gparallax_n$ and its uncertainty $\sigma_n$.
Importantly, to each \gaia\ parallax measurement we add a positive
offset $\varpi_0$ to adjust for the under-estimation of
parallaxes reported by the \gaia\ Collaboration (\citealt{lindegren}).
The \gaia\ recommended value is $\varpi_0=0.029$\,mas (\citealt{lindegren}),
but we adopt $\varpi_0 = 0.048$\,mas because that value
optimizes our cross-validation objective.
The spectrophotometric parallaxes do not depend strongly on this choice, except
for small differences at very small parallax (large distance).
The Training Sets are shown in \figurename~\ref{fig:training}.
\begin{figure}
\includegraphics[width=\textwidth]{./training_sample.pdf}
\caption{Color--color diagrams for the 28226 Training Set stars.
  Note that there is less color range in the Training Set than in the
  Parent Sample.\label{fig:training}}
\end{figure}

@@

For every star $n$ in the full Parent Sample we construct the feature
vector $x_n$ as
\begin{eqnarray}
x_n\T &\equiv& [1, G_n, \BP_n, \RP_n, J_n, H_n, K_n, W_{1n}, W_{2n}, \ln f_{1n}, \ln f_{2n}, \cdots, \ln f_{Ln}]
\end{eqnarray}
where the 1 permits a linear offset,
the photometry is from \gaia, \zmass, and \wise, respectively,
the fluxes in the $L=7405$ \apogee\ spectral pixels (for which there are reliably
and consistently data) for star $n$ are denoted
$f_{1n}$, $f_{2n}$, and so on,
and we have taken logarithms of those fluxes.
These feature vectors live in a $D$-dimensional space where $D=7414$.
For every star $n$ in either of the Training Sets we additionally require---in
addition to these feature vectors---a \gaia-measured astrometric parallax $\gparallax_n$
and uncertainty $\sigma_n$.

@@

This 9-percent precision estimate is conservative in the sense that it does not deconvolve
the contributions from the \gaia\ noise model.
However it is also not fully conservative, in that the test is performed with
\gaia's best stars, which are bright, nearby, in low extinction regions, and
(because of these things) near Solar metallicity.
There could in principle be additional bias or scatter for stars that are in dustier
regions or at lower metallicities.
\figurename~\ref{fig:residuals} shows that
there is no suggestion of any such trends when we color the residuals by metallicity
or $H-W_2$ color (which is a reddening proxy).
However, that figure does show that stars with greater reddening do appear to have
larger scatter against the \gaia\ parallaxes.
Additional tests confirm that our spectrophotometric parallaxes have largest scatter
in the dustiest and most crowded regions, as expected given the relatively low angular resolution
photometry.
\begin{figure}
\includegraphics[width=\textwidth]{./residuals_training.pdf}
\caption{Prediction residuals colored by relevant features.
  Each panel shows the prediction residuals (\gaia\ astrometric parallax minus
  spectrophotometric parallax $\gparallax_n-\sparallax_n$) colored by a different
  data feature that might be relevant to the residual.\label{fig:residuals}}
\end{figure}

@@

Another validation of the results can be obtained by looking at known
clusters or spatially compact objects in the data.
In \figurename~\ref{fig:clusters}, we show the parallax distribution from \gaia\ and
the spectrophotometric-parallax distribution from our model for stars that are
astrometrically confirmed members of known stellar clusters.
These clusters span some range in metallicity and age, and therefore test
the accuracy of the method for different kinds of populations.
With the exception of \acronym{M71}, these figures 
indicate that both \gaia\ and the spectrophotometric parallaxes appear to be unbiased
for these clusters and the range in abundances and ages they represent.
The case of cluster \acronym{M71} is troubling, but after searching for trends with
housekeeping data (also see \figurename~\ref{fig:residuals}) we don't find any correlates
with parallax offsets.
\begin{figure}
\includegraphics[width=\textwidth]{clusters.pdf}
\caption{The histograms of \gaia\ astrometric parallaxes $\gparallax_n$
  and spectrophotometric parallaxes $\sparallax_n$ for member stars for
  four stellar clusters in the \apogee\ observing footprint. The spectrophotometric
  parallaxes and their inverse-variance-weighted mean and the astrometric parallaxes
  and their inverse-variance-weighted mean
  the expected parallax of the cluster is shown as a vertical line.
  Cluster members were identified to be close to the cluster in celestial coordinates
  and comoving in proper motion.\label{fig:clusters}}
\end{figure}

@@

In the end, this method produces a spectrophotometric parallax estimate $\sparallax_m$ for
every star $m$.
\tablename~\ref{tab:output} shows a snippet of this output, with the full output available
as an electronic data table associated with this \documentname.
In order to use this output with \gaia\ astrometry, we recommend joining the table to the
\gaia\ Catalog on the \zmass\ ID (name), as per the instructions at the \gaia\ Archive (\citealt{gaiaarchive}).
In order to use this output with \apogee\ radial velocities and element abundances, we recommend
joining the table to the \sdssiv\ Catalog on the \zmass\ ID (name).
\begin{table}
\caption{HOGG TBD: The generated spectrophotometric parallaxes.\label{tab:output}}
\end{table}

The linear model permits straightforward propagation of uncertainty from the
feature inputs to the spectrophotometric parallaxes.
Although the model fundamentally assumes that the prediction features are noise-free
(see the ``good features'' assumption in \sectionname~\ref{sec:assumptions}),
there are in fact small uncertainties on the features that we can propagate to 
make uncertainty estimates on the spectrophotometric parallaxes.
We perform this propagation; these uncertainties are also given in
\tablename~\ref{tab:output}.

@@

One strong assumption of this work is that our set of features is sufficient to
build a predictive model for distance.
Our arguments for our included features is that they ought to be sufficient to
estimate a dust-independent apparent magnitude and the stellar parameters (especially
age and surface gravity) that are sensitive to luminosity.
In particular, since our model is linear in the logarithm of the parallax, it is
also linear in any kind of apparent magnitude at fixed distance and also linear
in distance modulus at fixed luminosity.
Our feature decisions were highly motivated by these kinds of considerations.
And the derived model parameters shown in \figurename~\ref{fig:spectral} do indicate that
the model is capturing some or all of the expected information in the spectra, at least.

>>>>

% To-do items
% -----------
% - Coryn's point that we should carefully explain how our model is geometric.
% - check with WISE experts that we have the correct WISE photometry
% - write words about other linear models we have made or could make, probably in the discusssion.
% - are we citing all the proper DR2 papers we need to cite?
% - HWR: Footnote to explain the use of the word ``feature'' for astronomers.
% - HWR: Figures 1 and 2 need to be combined to directly compare Parent and Training.
% - HWR: Figure 3: Note that the y-axis is positive-definite!
% - HWR: Figure 7: Caption needs to say what arrows represent.
% - Make the comparison with Cannon as a function of L1 strength and maybe look at correlation coefficients between parameter vectors.
% - Get model running on ASPCAP outputs and compare in Discussion.
% - Make sure we incorporate / respond to Ness comments from Aug 13 email. And thank Ness.
% - Thank APOGEE telecon and add paragraph about AGB stars and the color-magnitude diagram in the discussion.
% - Describe how we ran The Cannon; discussion of how we compare to The Cannon.
% - Cite and describe Eilers et al.
% - comment on parallax offset being a function of position and color
% - remove Cannon plot for the moment
% - comment on new plot of new precision on parallax compared to Gaia's precision
% - write caption and paragraph about the new figure showing how much better we are than Gaia at large distance
% - need a snippet for Table 1.

@@

% math macros
\newcommand{\T}{^{\mathsf{T}}}
\DeclareMathOperator*{\argmin}{argmin}
\newcommand{\logg}{\log g}
\newcommand{\BP}{{G_\mathrm{BP}}}
\newcommand{\RP}{{G_\mathrm{RP}}}
\newcommand{\gparallax}{{\varpi^{(\mathrm{a})}}}
\newcommand{\sparallax}{{\varpi^{(\mathrm{sp})}}}
\newcommand{\gsigma}{{\sigma^{(\mathrm{a})}}}
\newcommand{\ssigma}{{\sigma^{(\mathrm{sp})}}}

@@

\begin{abstract}\noindent
% context
With contemporary infrared spectroscopic surveys like \apogee,
red-giant stars can be observed to distances and extinctions
at which \gaia\ parallaxes are not highly informative.
Yet the combination of effective temperature, surface gravity, composition, and age---all
accessible through spectroscopy---determines a giant's luminosity.
Therefore spectroscopy plus photometry should enable 
precise spectrophotometric distance estimates.
% aims
Here we use the \apogee--\gaia--\zmass--\wise\ overlap to train a data-driven model
to predict parallaxes for red-giant branch stars with $0<\logg\leq2.2$.
% methods
We employ (the exponentiation of)
a linear function of \apogee\ spectral pixel intensities and multi-band photometry
to predict parallax spectrophotometrically.
The model training involves no logarithms or inverses of the \gaia\ parallaxes,
and needs no cut on the \gaia\ parallax signal-to-noise ratio.
It includes an L1 regularization to zero out the contributions of
uninformative pixels.
The training is performed
with leave-out subsamples such that no star's astrometry is used even indirectly in its
spectrophotometric parallax estimate.
% results
The model implicitly performs a reddening and extinction correction in its parallax prediction,
without any explicit dust model.
We assign to each star in the sample a new spectrophotometric parallax estimate;
these parallaxes have uncertainties of a few to 15 percent, depending on data quality,
which is more precise than the \gaia\ parallax for the vast majority of targets.
We can now get 10-percent distance estimates out to heliocentric distances of 20\,kpc,
and make global maps of the Milky Way's disk.
\end{abstract}

@@

If we want to make precise kinematic and element-abundance maps
of the Milky Way disk out to large heliocentric radii,
and especially through the extinction to the other side of the Galactic Center, we will
need to make use of luminous red giants, and we will need to observe in
the infrared.
These arguments motivate the \apogee\ (\citealt{apogee}),
\apogee-2, and \sdssv\ (\citealt{sdssv})
projects, which take spectra in the infrared, and deliver detailed abundances
along the entire red-giant branch up to the tip.
The maps made with these data will reveal the dynamics of the Milky Way and its disk,
and show us how and where stars form, and how they migrate around the
disk with time.

@@

Luminous red-giant stars are valuable for mapping the Milky Way disk for a number
of reasons, only one of which is their great luminosities (and hence brightnesses
even at large distance).
They are extremely common (frequent) stars, much more common than distance-indicating
stars like Cepheids.
They are produced by stellar populations at all metallicities and almost all ages,
so they can be found in all parts of the Galaxy.
They have temperatures and surface gravities such that it is
straightforward to spectroscopically measure metallicities,
detailed chemical abundances, and stellar parameters, even including ages (\citealt{martig, nessage}).
The red-giant branch is photometrically near-orthogonal to reddening vectors by dust,
so they are easy to de-redden or dust-correct.
And finally, in physical models of red giants, their predicted luminosities are simple
functions of composition, surface gravity, temperature, and age.
Thus, if we can measure these properties of red giants, we can (in principle) predict
their luminosities and use them as standard candles.
That is, red giants are standardizable candles.

@@

Data-driven models are very often more accurate than physical models of stars
for this kind of work, because they have the flexibility to discover
regularities or offsets in the data that are not adequately captured by the physical
models.
They do better than physical models when there are lots of training data
with properties that are good for performing regressions.

@@

\item[linearity] Perhaps the most restrictive assumption of this work is that
the logarithmic parallax (or, equivalently, the distance modulus) 
can be predicted as a \emph{completely linear function} of
the chosen features. We are only assuming this linearity in a small range of stellar
surface gravity, but this assumption is strong, and limits strongly the
flexibility or capacity of the model.
We make it to ensure that our method is easy to optimize, and the results are easy
to interpret.
It is also the case that linear models extrapolate better than higher-order or more
flexible models; this is not extremely important to the present work, but it does
protect our predictions at the edges (in terms of temperature, surface gravity,
and chemical abundances) of our sample.

\item[likelihood] We assume that the \gaia\ parallaxes are generated by a particular
stochastic process, in which the difference between the \gaia-Catalog parallax (plus
a small offset, which we learn below) and the true parallax is effectively drawn from a
Gaussian with a width set by the \gaia-Catalog uncertainty on the parallax.
This is the standard assumption in all properly probabilistic \gaia-Mission inferences
to date, but it subsumes a number of related assumptions, like that the \gaia\ noise
model is correct, that the stars are only covariant with one another at negligible levels, and that
there are no significant outliers or large-scale systematics.
\end{description}

@@

The model and log-likelihood function can be expressed heuristically as
\begin{eqnarray}
\gparallax_n &=& \exp(\theta\cdot x_n) + \mbox{noise}
\\
\chi^2(\theta) &\equiv& \sum_{n=1}^N \frac{[\gparallax_n - \exp(\theta\cdot x_n)]^2}{\gsigma_n^2}
\end{eqnarray}
where
$\gparallax_n$ is the \gaia\ measurement (adjusted; see below) of the astrometric parallax of star $n$,
the model is that the logarithm of the true parallax
can be expressed as a linear combination of the components
of some $D$-dimensional feature vector $x_n$,
$\theta$ is a $D$-vector of linear coefficients,
$\gsigma_n$ is the \gaia\ estimate of the uncertainty on the parallax measurement,
and $\chi^2(\theta)$ is (twice) the negative-log-likelihood for the parameters $\theta$
under the assumption of known Gaussian noise and
that there are $N$ independently measured stars $n$.
The feature vector $x_n$ contains photometry in a few bands, and all 7400-ish pixels
in the pseudo-continuum-normalized \apogee\ spectrum, so $D$ is on the order of 7400.

In addition, we assume that many entries in the parameter $D$-vector $\theta$ will be zero
(the ``sparsity'' assumption).
In order to represent this expectation,
we optimize not $\chi^2$ but a regularized objective function
\begin{eqnarray}
\hat{\theta} &\leftarrow& \argmin_{\theta}\left[\frac{1}{2}\,\chi^2(\theta) + \lambda\,||P\cdot\theta||_1^1\right]
\label{eq:objective}
\end{eqnarray}
where
$\lambda$ is a regularization parameter,
$P$ is a projection operator that selects out from the $\theta$ vector only those components
that pertain to the \apogee\ spectral pixels,
and $||y||_1^1$ is the L1-norm or sum of absolute values of the components of $y$.
This kind of regularization (L1) adds a convex term to the optimization and leads to
sparse optima.
The $P$ operator makes it such that we only regularize the components of $\theta$ that multiply
the spectral pixels; we ask for the spectral model to be sparse but we don't ask for the photometric
model to be sparse.
We set the value of $\lambda$ by cross-validation.

@@

Once the model is optimized, the output of the model is a prediction of the parallax,
or really what we will call the \emph{spectrophotometric parallax}.
This spectrophotometric parallax
$\sparallax_m$ for any star $m$ in the validation or test set is
assigned according to
\begin{eqnarray}
\sparallax_m &\leftarrow& \exp(\hat{\theta}\cdot x_m)
\end{eqnarray}
where
$\hat{\theta}$ is the optimal parameter vector according
to \equationname~(\ref{eq:objective}),
and
$x_m$ is the feature $D$-vector for star $m$.
The model can be trained on a training set of stars and applied to
a validation set or a test set to make predictions.
The only requirements are that every test-set object $m$  must have a full feature
vector $x_m$ just as every training-set object $n$ must have a full feature
vector $x_n$.

The linear model permits straightforward propagation of uncertainty from the
feature inputs to the spectrophotometric parallaxes.
Although the model fundamentally assumes that the prediction features are noise-free
(see the ``good features'' assumption in \sectionname~\ref{sec:assumptions}),
there are in fact small uncertainties on the features that we can propagate to 
make uncertainty estimates on the spectrophotometric parallaxes.
\begin{eqnarray}
\ssigma_m^2 &\leftarrow& \sparallax_m^2\,\hat{\theta}^T\cdot C_m\cdot\hat{\theta}
\end{eqnarray}
where $\ssigma_m$ is the uncertainty on the inferred (or predicted)
spectrophotometric parallax $\sparallax_m$,
the left-hand side is a scalar product,
and $C_m$ is the (presumed diagonal, actually) covariance matrix of the input
features.
That is, $C_m$ is a diagonal matrix with the uncertainty variances of the elements
of $x_m$ along the diagonal.

@@

We used the \gaia\ (\citealt{gaia}) \acronym{DR2} (\citealt{gaiadr2}) Data Archive (\citealt{gaiaarchive})
official matches (\citealt{xmatch}) to match the \apogee+\zmass\ stars
to the \wise\ Catalog (\citealt{wise}),
according to the \gaia\ Archive internal match criteria.
This gives, for each matching star $n$,
mid-infrared photometry $W_{1n}$ and $W_{2n}$ at 3.6 and 4.5\,$\mu$m
In detail we use the \code{w1mpro} and \code{w2mpro} Catalog entries.

@@

For training, we need A and B Training Sets.
We define the Training Set stars to be Parent-Sample stars that also
have a measured \gaia\ parallax.
The Training-Set stars, in addition to meeting all Parent-Sample cuts,
also must meet an uncertainty criterion $\gsigma_n < 0.1\,\mathrm{mas}$,
\begin{equation}
\code{parallax_error < 0.1}
\end{equation}
it must be observed by more than $8$ (widely separated) times
\begin{equation}
\code{visibility_periods_used >= 8}
\end{equation}
as well as meet the additional quality criterion
\begin{equation}
\code{astrometric_chi2_al / sqrt(astrometric_n_good_obs_al - 5) <= 35}
\end{equation}
which eliminates detections that are not consistent with a single PSF \citep[see][]{calj},
to ensure that the parallax measurements are good.
These cuts leave 28226 stars total in the Training Sets.
But---as we have emphasized above---we do not cut ever on parallax or
or the ratio to its uncertainty. For this reason, most of the Training-Set stars
do not have significantly measured parallaxes and more than 2~percent even have
negative parallaxes!
But we will perform our training such that it will be unbiased in these
circumstances.
Indeed it would be necessarily biased if ever we did cut on parallax or
parallax signal-to-noise.

In detail,
for each star $n$ in the Training Sets, we take from \gaia\ \acronym{DR2} (\citealt{gaiadr2})
the astrometric parallax $\gparallax_n$ and its uncertainty $\gsigma_n$.
Importantly, to each \gaia\ parallax measurement we add a positive
offset $\varpi_0$ to adjust for the under-estimation of
parallaxes reported by the \gaia\ Collaboration (\citealt{lindegren}).
The \gaia\ recommended value is $\varpi_0=0.029$\,mas (\citealt{lindegren}),
but we adopt $\varpi_0 = 0.048$\,mas because that value
optimizes our cross-validation objective.
HOGG CITE STASSUN MAYBE AND OTHERS?
The spectrophotometric parallaxes we generate
do not depend strongly on this choice, except
for small differences at very small parallax (large distance).
The Training Sets are shown in \figurename~\ref{fig:training}.
\begin{figure}
\includegraphics[width=\textwidth]{./training_sample.pdf}
\caption{Color--color diagrams for the 28226 Training Set stars.
  Note that there is less color range in the Training Set than in the
  Parent Sample.\label{fig:training}}
\end{figure}

@@

For every star $n$ in the full Parent Sample we construct the feature
vector $x_n$ as
\begin{eqnarray}
x_n\T &\equiv& [1, G_n, \BP_n, \RP_n, J_n, H_n, K_n, W_{1n}, W_{2n}, \ln f_{1n}, \ln f_{2n}, \cdots, \ln f_{Ln}]
\end{eqnarray}
where the 1 permits a linear offset,
the photometry is from \gaia, \zmass, and \wise, respectively,
the fluxes in the $L=7405$ \apogee\ spectral pixels (for which there are reliably
and consistently data) for star $n$ are denoted
$f_{1n}$, $f_{2n}$, and so on,
and we have taken logarithms of those fluxes.
These feature vectors live in a $D$-dimensional space where $D=7414$.
For every star $n$ in either of the Training Sets we additionally require---in
addition to these feature vectors---a \gaia-measured astrometric parallax $\gparallax_n$
and uncertainty $\gsigma_n$.

@@

This 9-percent precision estimate is conservative in the sense that it does not
deconvolve or correct for the contributions from the \gaia\ noise.
However it is also not fully conservative, in that the test is performed with
\gaia's best stars, which are bright, nearby, in low extinction regions, and
(because of these things) near Solar metallicity.
There could in principle be additional bias or scatter for stars that are in dustier
regions or at lower metallicities.
\figurename~\ref{fig:residuals} shows that
there is no suggestion of any such trends when we color the residuals by metallicity
or $H-W_2$ color (which is a reddening proxy).
However, that figure does show that stars with greater reddening do appear to have
larger scatter against the \gaia\ parallaxes.
Additional tests confirm that our spectrophotometric parallaxes have largest scatter
in the dustiest and most crowded regions, as expected given the relatively low angular resolution
photometry.
\begin{figure}
\includegraphics[width=\textwidth]{./residuals_training.pdf}
\caption{Prediction validation colored by relevant features.
  Each panel shows the predicted parallax \foreign{vs} the
  spectrophotometric parallax colored by a different
  data feature that might be relevant to the residual.\label{fig:residuals}}
\end{figure}

@@

Another validation of the results can be obtained by looking at known
clusters or spatially compact objects in the data.
In \figurename~\ref{fig:clusters}, we show the parallax distribution from \gaia\ and
the spectrophotometric-parallax distribution from our model for stars that are
astrometrically confirmed members of known stellar clusters.
These clusters span some range in metallicity and age, and therefore test
the accuracy of the method for different kinds of populations.
With the exception of \acronym{M71}, these figures 
indicate that both \gaia\ and the spectrophotometric parallaxes appear to be unbiased
for these clusters and the range in abundances and ages they represent.
The case of cluster \acronym{M71} is troubling, but after searching for trends with
housekeeping data (also see \figurename~\ref{fig:residuals}) we don't find any correlations
with parallax offsets.
\begin{figure}
\includegraphics[width=0.7\textwidth]{clusters.pdf}
\caption{The \gaia\ astrometric parallaxes $\gparallax_n$
  and spectrophotometric parallaxes $\sparallax_n$ for member stars for
  four stellar clusters in the \apogee\ observing footprint. The spectrophotometric
  parallaxes and their inverse-variance-weighted mean and the astrometric parallaxes
  and their inverse-variance-weighted mean
  are shown as vertical lines.
  Cluster members were identified to be close to the cluster in celestial coordinates
  and comoving in proper motion.\label{fig:clusters}}
\end{figure}

@@

In the end, this method produces a spectrophotometric parallax estimate $\sparallax_m$
and uncertainty $\ssigma_m$ for every star $m$.
\tablename~\ref{tab:output} shows a snippet of this output, with the full output available
as an electronic data table associated with this \documentname.
In order to use this output with \gaia\ astrometry, we recommend joining the table to the
\gaia\ Catalog on the \zmass\ ID (name), as per the instructions at the \gaia\ Archive (\citealt{gaiaarchive}).
In order to use this output with \apogee\ radial velocities and element abundances, we recommend
joining the table to the \sdssiv\ Catalog on the \zmass\ ID (name).
\begin{table}
\caption{HOGG TBD: The generated spectrophotometric parallaxes.\label{tab:output}}
\end{table}

@@

One strong assumption of this work is that our set of features is sufficient to
build a predictive model for distance.
Our argument for our included features is that they ought to be sufficient to
estimate a dust-independent apparent magnitude and the stellar parameters (especially
age and surface gravity) that are sensitive to luminosity.
In particular, since our model is linear in the logarithm of the parallax, it is
also linear in any kind of apparent magnitude at fixed distance and also linear
in distance modulus at fixed luminosity.
Our feature decisions were highly motivated by these kinds of considerations.
And the derived model parameters shown in \figurename~\ref{fig:spectral} do indicate that
the model is capturing some or all of the expected information in the spectra, at least.
